An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects in Time-Frequency Representations

### 2020-07-16 Notes/BoundEffRed.tex
paper

\section{Introduction}
\label{se:introduction}
\IEEEPARstart{I}{n} any digital acquisition system, the study and the interpretation of the measured signals generally require an analysis tool, which enables researchers to point out the useful characteristics of the signal. The need for signal analysis arises from various signals, ranging from audio~\cite{Stowell18computational,Muller11signal}, mechanical~\cite{Peng02vibration} or biomedical signals~\cite{Akay96detection}. In this paper, we focus on oscillatory time series. For instance, biomedical signal, such as photoplethysmogram (PPG), contain several characteristics, such as respiratory rate or blood pressure, that cannot be interpreted from its run-sequence plot in the time domain. An analysis tool would make possible the extraction of these useful characteristics.

Usually, the measured signals exhibit nonstationary behavior, and the observed quantities might be interfered by transient phenomena that can vary rapidly and irregularly. For example, the signal might oscillate fast with large amplitude at one moment, and then oscillate slow with small amplitude at the next moment. In order to adapt the analysis to nonstationarities, local spectral analysis is generally performed~\cite{Stoica05spectral,Matz97generalized}. The short time Fourier transform~\cite{Grochenig01foundations} (STFT) is a typical tool which is build for this purpose, and enables the determination of the local frequency content of a nonstationary signal.

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is is localized around the origin, and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently the results of the inner product~\eqref{eq:windowing} is distorted. This phenomenon is usually understood as the \emph{boundary effect}. We display on Fig.~\ref{fig:ex.intro} the result of the SST of a PPG (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on both sides of this representation. Indeed, while in the major central part of the image, clear lines stand out, they become blurred as they approach the left and right boundaries of the image. The right boundary is zoomed in and displayed on the right of Fig.~\ref{fig:ex.intro}, which emphasizes the result of boundary effects. Clearly, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window based TF analysis tools for the real-time analysis purpose. It is thus desirable to have a solution to eliminate this boundary effect.

@@

%The purpose of this paper is providing a fast algorithm allowing to tackle boundary effects and limit the disruptions it may cause on the signals representations based on windowing. 
Attempts to minimize the boundary effects generally consists in softening the discontinuity on signals edges. For instance, judicious choices of analysis windows whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane~\cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship of chosen analysis windows and TF analysis tool, these techniques do not make it possible to reduce the boundary effects of any TF representations, and for any analysis window. Another natural idea consists of carrying out a preliminary step of extending the signal beyond its boundaries,
%
and due to its flexibility, various extension schemes have been proposed. For example, there exists simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. There also exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD), the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR), the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}, and long short-term memory model \cite{vlachas2018data}. It is a far from exhaustive list, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. While the second class of extension schemes give better extended signals than the first class, they generally have a great computational cost. 
%

In this paper, we propose a fast extension algorithm based on a simple dynamical model, in order to optimize the trade-off between the extension quality and the computational cost. The proposed algorithm is composed of two steps.
\begin{enumerate}
\item \emph{Extend the signal by forecasting it.} The aim is to use a dynamic model to predict the values taken by the measured signal outside the measurement interval. Then, once this operation is done, we have access to an extended signal defined on a larger interval $I_\Delta$, where $\Delta$ denotes the size of the extension on both boundaries of $I$.
\item \emph{Run the local analysis tool on the extended signal.} Assuming that the support of the analysis window is smaller that $2\Delta$, the local analysis near the boundary of $I$ is now possible without lack of information thanks to knowledge brought by the extension on both sides. 
\end{enumerate}
Thus, assuming that the quality of the extension step is sufficient, the analysis results obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the non-extended signal. 
%
We claim and prove that forecasting oscillatory signals based on the simple dynamical model combined with the simple least square approach is sufficient for reducing the boundary effects for the TF analysis, or other kernel based analysis. See Figure \ref{fig:ex.intro} for a snapshot of the result. The main benefit of this simple approach is a numerically efficient solution with theoretical guarantee for the real-time analysis purpose.

The paper is organized in the following. In section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In section~\ref{se:theoretical}, we show that, even though we consider a simplistic dynamic model, it is sufficient to extend signals takings the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on a class of signals, the sums of sine waves, is given in section~\ref{se:theoretical}. In section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals such as respiratory signals, and compare it to the theoretical results.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the reduction of boundary effects on TF representations in real-time. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient for the extension of signals formed by a sum of sine waves. Moreover, the numerical results show that this strategy allows us to switch to a real-time implementation of {\sf BoundEffRed} is possible, due to the low computation time, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many time-frequency representations. 
%Other numerical examples are also included, such as the application of {\sf BoundEffRed} to other types of biomedical signals such as ECGs.

Various improvements can be considered to make the algorithm more robust. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting is no longer relevant, and the TF representations are not ideal. A preliminary step should then be added to the algorithm to detect the portions to be rejected, for which a better forecasting scheme is needed. A more fundamental improvement would be to perform the prediction step directly in the TF plane, or even with the aim of specifically predicting the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.

>>>>

\section{Introduction}
\label{se:introduction}
\IEEEPARstart{I}{n} any digital acquisition system, the study and the interpretation of the measured signals generally require an analysis tool, which enables researchers to point out the useful characteristics of the signal. The need for signal analysis arises from various signals, ranging from audio~\cite{Stowell18computational,Muller11signal}, mechanical~\cite{Peng02vibration}, or biomedical signals~\cite{Akay96detection}. For instance, biomedical signals, such as photoplethysmogram (PPG), contain several characteristics, including respiratory rate or blood pressure, that cannot be interpreted from its run-sequence plot in the time domain. An analysis tool would make possible the extraction of these useful characteristics.

Usually, the measured signals exhibit nonstationary behavior, and the observed quantities might be interfered by transient phenomena that can vary rapidly and irregularly. In this paper, we focus on oscillatory time series. These signals might, for example, oscillate fast with large amplitude at one moment, and then oscillate slowly with small amplitude at the next moment. In order to adapt the analysis to nonstationarities, local spectral analysis is generally performed~\cite{Stoica05spectral,Matz97generalized}. The short-time Fourier transform~\cite{Grochenig01foundations} (STFT), a typical tool built for this purpose, enables the determination of the local frequency content of a nonstationary signal.

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is localized around the origin and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently, the results of the inner product~\eqref{eq:windowing} are distorted. This phenomenon is usually understood as the \emph{boundary effect}. We display on the bottom left of Fig.~\ref{fig:ex.intro} the result of the SST of a PPG (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on the right side of this representation. Indeed, while in the major left part of the image, clear lines stand out, they become blurred as they approach the right boundary of the image. Therefore, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window-based TF analysis tools for real-time analysis purposes. It is thus desirable to have a solution to eliminating the boundary effects.

@@

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. For instance, judicious choices of analysis windows whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane~\cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship of chosen analysis windows and TF analysis tool, these techniques do not make it possible to reduce the boundary effects of any TF representations, and for any analysis window. Another natural idea consists of carrying out a preliminary step of extending the signal beyond its boundaries, and due to its flexibility, various extension schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. There also exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD), the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR), the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}, and long short-term memory model \cite{vlachas2018data}. It is a far from an exhaustive list, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. While the second class of extension schemes gives better-extended signals than the first class, they generally have a great computational cost.

In this paper, we propose a fast extension algorithm based on a simple dynamical model, in order to optimize the trade-off between the extension quality and the computational cost. The proposed algorithm is then composed of two steps.
\begin{enumerate}
\item \emph{Extend the signal by forecasting it.} The aim is to use a dynamic model to predict the values taken by the measured signal outside the measurement interval. Then, once this operation is done, we have access to an extended signal defined on a larger interval $I_\Delta$, where $\Delta$ denotes the size of the extension on both boundaries of $I$.
\item \emph{Run the local analysis tool on the extended signal.} Assuming that the support of the analysis window is smaller than $2\Delta$, the local analysis near the boundary of $I$ is now possible without lack of information thanks to knowledge brought by the extension. 
\end{enumerate}
Thus, assuming that the quality of the extension step is sufficient, the analysis results obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the non-extended signal. We claim and prove that forecasting oscillatory signals based on the simple dynamical model combined with the simple least square approach is sufficient for reducing the boundary effects for the TF analysis, or other kernel-based analysis. See the bottom right of Fig.~\ref{fig:ex.intro} for a snapshot of the result. The main benefit of this simple approach is a numerically efficient solution with a theoretical guarantee for real-time analysis purposes.

The paper is organized in the following. In section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In section~\ref{se:theoretical}, we show that the dynamic model we consider is sufficient to extend signals takings the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on a class of signals, the sums of sine waves, is given in section~\ref{se:theoretical}. In section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals, such as respiratory signals, and compare it to the theoretical results.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the real-time reduction of boundary effects in TF representations. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient to extend signals formed by a sum of sine waves. Moreover, thanks to the low computational time, the numerical results show that this strategy allows us to switch to a real-time implementation of {\sf BoundEffRed}, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many time-frequency representations. 
%Other numerical examples are also included, such as the application of {\sf BoundEffRed} to other types of biomedical signals such as ECGs.

Various improvements can be considered to make the algorithm more robust. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting step is no longer relevant, and only slows down the calculation of the TF representation. A preliminary step should then be added to the algorithm to detect signal activity, and disable the forecasting step when possible. A more fundamental improvement would be to perform the forecasting step directly in the TF plane, to specifically predict the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.
